Leadership across cultures and contexts: an exploratory analysis

### 2020-05-15 leadership_across_cultures_contexts.Rmd
Minor revisions throughout, grammatical, copy-edit.

<!-- CLOSING -->
We first briefly review the qualities of leaders, the functions they serve, and the costs and benefits of leadership across nonindustrial non-WEIRD human societies. We then outline the aims and methods of the current cross-cultural study -- which focuses on exploring patterns among non-WEIRD populations -- before presenting results and discussion.

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009, drawing on the ethnography of highland New Guinea, proposes that leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling military formidability to rival groups. Among many ethnographically described small-scale societies, which generally lack institutionalized political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when group demands require a managerial presence [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Leaders also play an important role in ritual contexts, such as in initiations [@singh_cultural_2017], facilitating marriages [@walker_evolutionary_2011] and organizing feasts [@wiessner_food_1996]. Between-group interactions, including both conflict and cooperation, are also common among hunter-gatherers and other nonindustrial societies [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Representing the group and managing internal and external relationships and conflicts are expected to be the most salient functions of smaller-scale community leaders.

@@

Evolutionary theoretical models and studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and contexts and the functions leaders serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others for generalizations [e.g., @garfield_review_2019; @von_rueden_leadership_2015]. Here, we seek to provide the systematic and near comprehensive view of the qualities and functions of leaders across human societies from a representative cross-cultural sample of nonindustrial, non-WEIRD populations. We also provide the first comparative view describing the costs and benefits of leadership for both leaders and followers in an evolutionary biological framework. We also aim to explore how these dimensions of leadership vary within populations across social contexts and across populations with different subsistence strategies, as well as by leader sex and continental region.

@@

*Coercive authority* was documented in about 57% of cultures. Unlike other more frequent leader qualities, which were generally positive or prosocial traits, *Coercive authority* was the most frequent negative or dominance-style leaders quality. We therefore chose to perform additional exploratory analyses with *Coercive authority* as the outcome in logistic elasticnet regression models similarly as described above for *High status* (see Figure \@ref(fig:elasticcoercive) caption for details).

@@

Words that were strong predictors of leaders with coercive authority included, *power* and *authority* (unsurprisingly), as well as *territory*, *chief*, and *control*; the word *leadership* was the only strong negative predictor. See Figure \@ref(fig:elasticcoercive)A. Using all other quality and function variables as predictors identified *Punishment* as the strongest predictor of evidence for *Coercive authority*. The qualities of *Aggressiveness* and *Feared* were also positive predictors, as were the functions, *Political appointments*, *Control economics*, and *Control immigration*. Furthermore, evidence for leaders lacking coercive authority (our *No coercive authority* variable) was documented in `r nocoerciveauthority`% of cultures (Figure \@ref(fig:varsupport)) and demonstrated variability by subsistence strategy with hunter-gatherers providing relatively higher levels of evidence for leaders lacking coercive authority. See Figure S\@ref(fig:heatmapdrop1).

@@

>>>>

<!-- CLOSING -->
We first briefly review the qualities of leaders, the functions they serve, and the costs and benefits of leadership across nonindustrial, non-WEIRD human societies. We then outline the aims and methods of the current cross-cultural study -- which focuses on exploring patterns among non-WEIRD populations -- before presenting results and discussion.

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009, drawing on the ethnography of highland New Guinea, proposes that leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling military formidability to rival groups. Among many ethnographically described small-scale societies, which generally lack institutionalized political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when group demands require a managerial presence [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Leaders also play important roles in ritual contexts, such as in initiations [@singh_cultural_2017], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Between-group interactions, including both conflict and cooperation, are also common among hunter-gatherers and other nonindustrial societies [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Representing the group and managing internal and external relationships and conflicts are expected to be the most salient functions of smaller-scale community leaders.

@@

Evolutionary theoretical models and studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and contexts and the functions leaders serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others for generalizations [e.g., @garfield_review_2019; @von_rueden_leadership_2015]. We seek to provide a systematic and near comprehensive view of the qualities and functions of leaders across human societies from a representative cross-cultural sample of nonindustrial, non-WEIRD populations. We also provide the first comparative view describing the costs and benefits of leadership for both leaders and followers in an evolutionary biological framework. We also aim to explore how these dimensions of leadership vary within populations across social contexts and across populations with different subsistence strategies, as well as by leader sex and continental region.

@@

*Coercive authority* was documented in about 57% of cultures. Unlike other more frequent leader qualities, which were generally positive or prosocial traits, *Coercive authority* was the most frequent negative or dominance-style leader quality. We therefore chose to perform additional exploratory analyses with *Coercive authority* as the outcome in logistic elasticnet regression models similarly as described above for *High status* (see Figure \@ref(fig:elasticcoercive) caption for details).

@@

Words that were strong predictors of leaders with coercive authority included, *power* and *authority* (unsurprisingly), as well as *territory*, *chief*, and *control*; the word *leadership* was the only strong negative predictor. See Figure \@ref(fig:elasticcoercive)A. Using all other quality and function variables as predictors identified *Punishment* as the strongest predictor of evidence for *Coercive authority*. The qualities of *Aggressiveness* and *Feared* were also positive predictors, as were the functions, *Political appointments* and *Control economics*. Furthermore, evidence for leaders lacking coercive authority (our *No coercive authority* variable) was documented in `r nocoerciveauthority`% of cultures (Figure \@ref(fig:varsupport)) and demonstrated variability by subsistence strategy with hunter-gatherers providing relatively higher levels of evidence for leaders lacking coercive authority. See Figure S\@ref(fig:heatmapdrop1).

@@

Below we briefly review leadership functions and qualities in WEIRD contexts, to suppliment our discussions of non-WEIRD leaderhsip dynamics in the main text.
